Who Gets Cared For and Why? Fertility Shocks, Intra-Household Investment Reallocation, and the Quantity--Quality Trade-Off in the Philippines

### 2025-08-07 docs/paper.tex
revise paper title, abstract, intro, and literature review
Updated the paper's title, short title, acknowledgements, and abstract to reflect a new focus on fertility shocks and intra-household investment in the Philippines. Substantially revised the introduction and literature review for clarity, conciseness, and to better align with the new research questions. Added new LaTeX header includes and made minor technical corrections throughout the RMarkdown file. Added a new preamble.tex file and updated references.

@@

\title{Womb for One More? Fertility Expansion and the Limits of Early
Health Investment: Empirical Evidence on the Quantity-Quality Trade-off
from a Contraceptive Ban in Urban Philippines}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

\author{
  Erika Salvador\\
  Caroline Theoharides\thanks{
  Salvador: Amherst
College, \href{mailto:esalvador28@amherst.edu}{esalvador28@amherst.edu}.
  Theoharides: Amherst
College, \href{mailto:ctheoharides@amherst.edu}{ctheoharides@amherst.edu}.
  This research was made possible by the Schupf Fellowship at Amherst
  College. I am indebted to my faculty adviser, Professor Caroline
  Theoharides, for her mentorship and support throughout the development
  of this project. I also wish to thank Faculty Director Professor
  Amelie Hastie and the campus partners whose institutional support
  enabled this work. I am further grateful to the Departments of
  Economics and Mathematics \& Statistics for their academic support.
  The views expressed and any errors contained herein are entirely my
  own responsibility.
}
}

\date{\today}
\pubMonth{07}
\pubYear{2025}
\pubVolume{}
\pubIssue{}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

\begin{abstract}
This paper examines whether increased fertility, driven by reduced
access to modern contraception, affects early-life health outcomes in
low-income urban households. I exploit a natural experiment created by a
2000 policy in Manila, Philippines---Executive Order No.~003---which
effectively banned the provision of modern contraceptives in all
city-run health facilities. Using a difference-in-differences design and
nationally representative data from the Philippine Demographic and
Health Surveys, I estimate the policy's impact on child health by
comparing outcomes in Manila before and after the ban to those in
comparable urban areas unaffected by the policy. The outcomes of
interest include birth weight (in grams), the mother's perceived size of
the child at birth, and receipt of key childhood vaccinations.
\end{abstract}

@@

Empirical investigations of the quantity-quality trade-off have focused
primarily on educational outcomes. Most studies examine how increases in
maternal fertility---or equivalently, the number of siblings a child
has---affect school enrollment, grade progression, test scores, and
completed years of schooling
\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other dimensions of child quality--such as health at birth and early
childhood health investments---are early-onset or biologically
constrained, and they tend to be less amenable to remediation later in
life. For instance, babies born with very low birth weight face higher
risks of infant mortality and long-term developmental deficits that
cannot be fully reversed \citep{cook2015understanding}. Similarly, a
child who misses critical early vaccinations is more vulnerable to
infectious diseases that can kill or disable them, which results to
lifelong negative health impacts \citep{costa2024child}. Such early-life
health setbacks may permanently impair cognitive development and
physical growth. This limits long-run productivity regardless of
subsequent improvements in schooling or incom

@@

I use this natural experiment to estimate the causal impact of increased
family size on early-childhood health outcomes. The analysis uses
nationally representative data from multiple rounds of the Philippine
Demographic and Health Survey (DHS), which include detailed information
on household structure, maternal characteristics, and child health
indicators. The primary outcomes are birth weight (measured in grams,
when reported) and the mother's subjective assessment of the child's
size at birth---both of which serve as indicators of neonatal health
status. In addition, I assess whether the child received key early-life
vaccinations, including those against polio,
diphtheria--pertussis--tetanus (DPT), and measles, as a measure of
essential postnatal health investment. These outcomes are widely
recognized indicators of child well-being and reflect both biological
and behavioral dimensions of parental investment in early development.
The empirical strategy follows a difference-in-differences design that
compares child outcomes in Manila and comparable urban areas before and
after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the
absence of the contraceptive ban, trends in child health outcomes in
Manila would have evolved in parallel with those in the comparison
cities. Second, no other policy changes or economic shocks in Manila
during the study period coincided with the timing and scope of the
contraceptive policy. I probe these assumptions using falsification
tests, placebo comparisons, and robustness checks that include
city-specific time trends, mother fixed effects, and controls for
baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2019sex} to confirm that the contraceptive ban led to an
increase in fertility among affected women. Next, I estimate the
reduced-form effects of exposure to the policy on the selected child
health outcomes (birth weight, perceived birth size, and vaccination
uptake). Finally, I examine heterogeneity in these effects across
subgroups defined by maternal education, household wealth, and access to
prenatal care. These characteristics proxy for a household's resources
and its capacity to buffer the impact of increased family size on child
health.

@@

In what follows, I review the theoretical foundations of the Q-Q model
and its evolution in the literature. I begin with the static models of
Becker \citep{becker1960economic} and Becker--Lewis
\citep{becker1973interaction}, which first formalized the trade-off
within a household utility maximization framework. We then examine
extensions to dynamic, intergenerational settings, including the
contributions of Becker and Tomes \citep{becker1976child} on child
endowments and the altruistic dynastic model associated with Barro and
Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and
unified growth models, notably Galor and Weil
\citep{galor2000population} and Galor and Moav \citep{galor2002natural},
which integrate the Q--Q mechanism into a general theory of demographic
and economic transformation.

@@

Beyond survival, early-life health indicators such as birth weight,
perceived size at birth, and vaccination status are important dimensions
of child quality, especially in low-income settings. When resources are
limited, parents may struggle to meet the basic needs of each child.
Empirical evidence aligns with the quantity--quality framework: children
from larger families often have worse health outcomes at birth and are
less likely to receive core immunizations. For example, recent
multi-country data show that the share of completely unimmunized
children---so-called ``zero-dose'' children---increases from 10.5\%
among those with no siblings to 17.2\% among those with four or more
\citep{costa2024child}. Studies from Southeast Asia also report that
higher birth order correlates with lower average birth weight and
smaller perceived size at birth. These patterns suggest that as
fertility increases, the health of each child may suffer due to more
limited parental investment. Studies from Southeast Asia
\citep{chen2021population} also report that higher birth order
correlates with lower average birth weight and smaller perceived size at
birth

@@

The results are striking. The contraceptives ban led to a significant
rise in births and family size in Manila city relative to the control
areas. Correspondingly, children born in Manila during the ban era
experienced a sizable decline in educational attainment -- clear
evidence of a Q--Q trade-off precipitated by the shock. In the authors'
words, the policy-driven increase in family size ``provide{[}s{]}
evidence of a quality--quantity trade-off'': larger families, forced by
the ban, resulted in lower schooling outcomes per child. By leveraging
an actual anti-contraception policy, this study offers causal
confirmation that constraints on family planning can worsen child
quality. Together with Orbeta's findings, it reinforces the relevance of
Becker's hypothesis in the Philippine setting: whether via unintended
fertility (Manila's case) or ordinary variation in sibling sex mix, more
children mean fewer resources per child and thus worse education
outcomes. Consistent with theory, Filipino parents facing exogenous
increases in quantity were unable to maintain the same level of quality
per child.

@@

This study builds upon the foundational framework of the
quantity-quality (Q--Q) trade-off, first articulated in
\citet{becker1960economic} and later refined in
\citet{becker1973interaction}. In this model, fertility and child
quality are jointly determined under a full-income constraint, such that
increases in the number of children---without a corresponding expansion
in household resources---necessitate a reduction in the average
investment per child. While empirical applications of this framework
have traditionally focused on educational outcomes, child quality can
also be expressed in terms of early-life health, including indicators
that emerge at birth or shortly after, such as birth weight, perceived
size at birth, and uptake of essential vaccinations.

@@

In the presence of a fertility shock, holding \(I\) constant, the
average \(\kappa_j\) must fall unless offset by changes in parental
behavior or external transfers. If these inputs decline and are critical
to early-life development, observable deficits may arise in birth weight
or perceived neonatal size, and children may be less likely to receive
complete immunization schedules.

The first hypothesis derived from this framework is that an exogenous
increase in fertility leads to a decline in per-child health investment,
particularly in low-income urban households with little financial slack.
A second, related hypothesis posits that this decline results in
measurable early-life health disadvantages---specifically, lower birth
weight, smaller perceived size at birth, and reduced uptake of key
childhood vaccinations. These hypotheses describe a causal pathway
through which fertility shocks may contribute to early deprivation and
potentially reinforce intergenerational cycles of poor health and
limited human capital.

@@

The primary data source for this study is the Philippine Demographic and
Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
These nationally representative surveys provide detailed child- and
household-level information, including maternal characteristics,
fertility behavior, and early-childhood health outcomes. Importantly,
they are the only Philippine DHS waves with available GPS cluster data,
which I use to approximate geographic location and assign respondents to
treatment and control groups. Because city-level identifiers are not
released in the public-use files, I rely on coarse geographic assignment
based on cluster coordinates. Clusters located within or near the
administrative boundaries of the City of Manila are classified as
``treated,'' while clusters located in other parts of the National
Capital Region (NCR) serve as the control group. In robustness checks, I
restrict the control group to Quezon City or expand it to the rest of
NCR excluding both Manila and Quezon City.

The analytic sample includes children under five years of age living
with their mothers at the time of the survey. I focus on three
early-life health outcomes that reflect both biological endowments and
parental investment: (1) birth weight, reported in grams when available;
(2) the mother's subjective assessment of birth size; and (3)
vaccination uptake, defined as receipt of key early-childhood
immunizations, including BCG, DPT, polio, and measles. These outcomes
are consistently available across all three waves and serve as proxies
for child quality in the context of the quantity--quality trade-off.

The 1998 DHS is used as the primary pre-policy baseline wave because it
captures children born before the implementation of Manila's
contraceptive ban in 2000. The 2008 DHS serves as the post-policy round
to capture outcomes for children born well after the ban took effect. I
also include the 2003 DHS in the main specification, recognizing that
children in this round may have been born shortly before or after the
policy's introduction. I treat 2003 as a transition-period dataset and
conduct sensitivity checks to assess the robustness of findings to its
inclusion. I exclude the 1993 DHS from the analysis due to the absence
of GPS data, which makes it impossible to reliably assign treatment
status.

To support the identification strategy, I use DHS summary indicators on
contraceptive use disaggregated by region and city. These data confirm a
sharp and sustained decline in modern contraceptive use in Manila
following the implementation of Executive Order No.~003, relative to
surrounding cities, validating the policy as a plausibly exogenous
fertility shock. I also considered using data from the Philippine
National Nutrition Survey (NNS), which includes rich anthropometric and
dietary data. However, the NNS microdata are only available from 2003
onward, which limits their value for pre-policy analysis. As such, the
DHS remains the most suitable dataset for this study.

\subsection{Identification Strategy}

To estimate the causal effect of increased fertility on early-childhood
health outcomes, I exploit a policy shock generated by Executive Order
No.~003, issued in 2000 by the mayor of Manila. This order restricted
access to modern contraceptives in all city-run health facilities and
remained in effect for nearly a decade. The policy was implemented only
in the City of Manila and did not affect neighboring municipalities
within the National Capital Region (NCR), which continued to provide
modern family planning services. This spatially localized and time-bound
restriction created a natural experiment that allows for a
quasi-experimental research design. I define the treatment group as
children residing in GPS clusters located within or near the Manila
boundary, and the control group as children in other NCR municipalities.
Given the limitations of the DHS GPS data, which are displaced for
confidentiality and do not include exact city identifiers, I use a
coarser geographic classification to approximate exposure. Robustness
checks vary the control group to include Quezon City only, or the rest
of NCR excluding both Manila and Quezon City.

The core of my empirical strategy is a difference-in-differences (DiD)
design that compares changes in child outcomes in Manila before and
after the ban relative to changes in similar urban areas that were
unaffected. The key assumption is that, in the absence of the policy,
trends in early-childhood health outcomes would have followed similar
trajectories across treatment and control areas. This approach accounts
for both fixed differences between cities and nationwide time trends.

To address potential violations of the parallel trends assumption---such
as Manila-specific shocks that may have affected all children regardless
of family size---I implement a triple-differences strategy that
leverages variation in exposure by maternal age. Because fertility
declines sharply with age, older women were biologically less likely to
respond to the contraceptive ban. For example, mothers aged 45--49 at
the time of the policy were near the end of their reproductive years and
thus unlikely to experience a fertility increase. By comparing changes
in child outcomes across maternal age cohorts within Manila and control
cities, I can account for time-varying, location-specific confounders
that affect all families, while isolating the policy's effect on family
size and its consequences.

I further restrict my main analysis to children born before the policy
was implemented. Although these children were not conceived due to the
ban, they may still have been indirectly affected by the fertility
increase through resource dilution within households. This strategy
avoids selection bias related to unobservable differences in the
endowments of ``marginal children'' born because of the policy. As
argued by Rosenzweig and Zhang (2009), when parents reinforce rather
than compensate for birth endowments, estimates based on older siblings
provide a lower bound of the true effect of fertility on child quality.
Empirical evidence from the Philippines supports this pattern of
reinforcement, as shown by Baguet and Dumas (2015). Taken together,
these strategies allow me to estimate the causal impact of fertility
shocks on early-life health while addressing key concerns about internal
validity and sample composition.

\subsection{Sample Selection}

The analytic sample is drawn from the Philippine Demographic and Health
Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003,
and 2008. I restrict the sample to children under age five (0--59
months) with non-missing information on key outcomes: birth weight (in
grams, when available), perceived size at birth, and vaccination status.
Because the Manila contraceptive ban was enacted in 2000 and likely
implemented unevenly at first, I define children born in 2001 and later
as post-policy and those born before 1997 as pre-policy. Children born
between 1997--2000 are excluded in the main specification but retained
in robustness checks.

I include all live births regardless of birth order or plurality to
preserve variation in sibship size. Twins and higher-order multiple
births are flagged for sensitivity checks but not excluded. Each child
is linked to their biological mother using maternal identifiers,
allowing construction of sibship size, birth order, and maternal age.
Exposure to the policy is defined geographically: children are
considered treated if they were born in the City of Manila, and
untreated if born in other municipalities within the National Capital
Region (NCR). Because GPS data are coarse in some years, geographic
exposure is defined at the city level rather than using precise
coordinates. Households outside NCR are excluded from the main analysis
to ensure urban comparability and avoid confounding by rural fertility
and health dynamics.

@@

The primary outcome variables reflect early-life health and investment.
These include: (1) birth weight in grams, as recorded by the mother when
available; (2) the mother's subjective assessment of birth size
(categorized as very small, small, average, large, or very large); and
(3) indicators for whether the child received core early-life
vaccinations, including BCG, DPT1--3, polio doses, and measles. A binary
variable indicating whether the child received all recommended vaccines
is also constructed as a summary measure of postnatal health investment.

The main explanatory variable is sibship size, defined as the total
number of surviving children ever born to the mother at the time of the
child's birth, as reported in the birth history module. This variable
captures the resource dilution mechanism central to the
quantity--quality trade-off. To assess whether the contraceptive ban
caused an increase in sibship size, I also construct an indicator for
policy exposure, coded as 1 if the child was born in the City of Manila
and in or after 2001 (i.e., post-ban), and 0 otherwise. In extended
specifications, I interact this treatment indicator with maternal age at
birth to capture heterogeneity in fertility effects across age cohorts.

Control variables include standard child-, mother-, and household-level
characteristics. At the child level, I include sex, birth order, whether
the child is part of a multiple birth, and the preceding birth interval
in months. At the maternal level, I control for the mother's age at
birth, educational attainment (categorized as none, primary, secondary,
higher), literacy status, employment status, and age at first birth.
Household-level controls include household size, urban/rural status
(restricted to NCR in the main sample), access to improved water and
sanitation, electricity, household crowding, and the DHS wealth index.
Fixed effects for survey year are included to account for common shocks
and secular trends. All regressions are weighted using DHS sampling
weights, and standard errors are clustered at the PSU (primary sampling
unit) level.

@@

To estimate the causal effect of fertility on early-life health
outcomes, I use a two-stage least squares (2SLS) strategy. Specifically,
I regress child health outcomes, such as birth weight, perceived size at
birth, and immunization uptake, on measures of sibship size, using
exogenous variation induced by the contraceptive supply ban in Manila as
an instrument. This approach allows me to isolate the impact of
increased fertility on child quality and mitigate bias from endogenous
family size decisions.

The estimation proceeds as follows:

\[
\text{First stage:} \quad K_{ijt} = \alpha_2 X_{ijt} + \gamma Z_{ijt} + v_{ijt}
\]

\[
\text{Second stage:} \quad Y_{ijt} = \alpha_1 X_{ijt} + \beta \widehat{K}_{ijt} + u_{ijt}
\]

where: \(Y_{ijt}\) is the health-related outcome for child \emph{i} in
city \emph{j} and survey year \emph{t}, including birth weight (grams),
perceived size at birth, and full basic immunization status. \(K_{ijt}\)
is sibship size at time of birth, defined as the number of surviving
siblings. \(Z_{ijt}\) is the instrument for sibship size, based on
exposure to the contraception ban---specifically, an interaction of a
Manila indicator and a post-ban cohort indicator. \(X_{ijt}\) is a
vector of control variables including child-, maternal-, and
household-level covariates (see \emph{Variables} section). Standard
errors are clustered at the PSU level; survey-year fixed effects are
included.

To account for heterogeneity in fertility responses to the ban, I also
estimate an alternative specification where the instrument interacts
treatment status (Manila × post-ban) with maternal age at birth. This
reflects the fact that older women were less likely to change fertility
behavior due to lower fecundity, and strengthens identification through
cross-cohort variation in exposure intensity.

Although the outcomes are binary or ordinal, I estimate equation (1)
using linear probability models to retain consistency with the 2SLS
framework. As robustness, I also estimate probit and logit models and
find qualitatively similar results. All models use DHS sampling weights,
and standard errors are clustered by primary sampling unit (PSU).

\section{Results}

\bibliographystyle{aea}
\bibliography{references}

\% The appendix command is issued once, prior to all appendices, if any.
\appendix

\section{Mathematical Appendix}

>>>>

\usepackage{pdfpages}
\usepackage{float}
\usepackage{makecell}
\setlength{\parskip}{0pt}       % Remove vertical space between paragraphs
\setlength{\parindent}{15pt}    % Indent first line of each paragraph
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{siunitx}

    \newcolumntype{d}{S[
      table-align-text-before=false,
      table-align-text-after=false,
      input-symbols={-,\*+()}
    ]}

@@

\title{Who Gets Cared For and Why? Fertility Shocks, Intra-Household
Investment Reallocation, and the Quantity--Quality Trade-Off in the
Philippines}
\shortTitle{Spillovers from Policy-Induced Fertility}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

\author{
  Erika Salvador\\
  Caroline Theoharides\thanks{
  Salvador: Amherst
College, \href{mailto:esalvador28@amherst.edu}{esalvador28@amherst.edu}.
  Theoharides: Amherst
College, \href{mailto:ctheoharides@amherst.edu}{ctheoharides@amherst.edu}.
  This research was made possible through the support of the Schupf
  Fellowship at Amherst College. I am deeply thankful to Dr.~Axel Schupf
  '57 for his generosity and vision in establishing this program, which
  annually supports a cohort of twenty rising sophomores and juniors in
  independent summer research. I feel truly privileged to have been
  selected through this competitive process and to have spent eight
  weeks immersed in a project that both challenged and inspired me. I am
  especially grateful to my mentor, Prof.~Caroline Theoharides, for her
  mentorship throughout this project---and more broadly, for sparking my
  interest in economic research and its potential to serve the public
  good in the Philippines. I also wish to thank Faculty Director
  Professor Amelie Hastie and the many campus partners whose
  institutional support made this experience possible. Finally, I am
  grateful to the Departments of Economics and Mathematics \& Statistics
  for equipping me with the analytical and conceptual tools that made
  this research possible---and for preparing me to take on independent
  research with confidence and curiosity. All views and any errors are
  my own.
}
}

\date{\today}
\pubMonth{08}
\pubYear{2025}
\pubVolume{}
\pubIssue{}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

\begin{abstract}
This paper investigates how increases in fertility, driven by reduced
access to contraception, affect parental investment in child health. I
use Executive Order No.~003---a 2000 policy in Manila that restricted
access to modern contraceptives---as an exogenous fertility shock.
Drawing on pooled data from the 1998, 2003, and 2008 Philippine
Demographic and Health Surveys, I estimate two-stage least squares
models that use policy exposure to instrument for sibship size. Child
health inputs fall into two categories: prenatal (proxied by birth
weight) and postnatal (measured by the total number of vaccine doses
received by age five). The first-stage results indicate that affected
mothers had, on average, 0.20 more living children at the time of a
birth. In the second stage, each additional sibling is associated with a
reduction of 22.97 vaccine doses received by the index child, a decline
large enough to prevent completion of the recommended immunization
schedule. The estimated effect on birth weight is −8,676 grams, though
statistically insignificant Reduced-form estimates are directionally
consistent: policy exposure is associated with a 0.91-dose drop in
vaccine uptake and a 349.60-gram decrease in birth weight. While
instrumental variable estimates are limited by a weak first stage, the
magnitude and sign of the coefficients support the hypothesis that
larger sibship sizes reduce health investments per child. The results
suggest that postnatal investments, particularly those requiring
sustained parental effort, are more sensitive to fertility shocks than
prenatal outcomes.
\end{abstract}

@@

Empirical investigations of the quantity-quality trade-off have focused
primarily on educational outcomes. Most studies examine how increases in
maternal fertility---or equivalently, the number of siblings a child
has---affect school enrollment, grade progression, test scores, and
completed years of schooling
\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other dimensions of child quality--such as health at birth and early
childhood health investments---are early-onset or biologically
constrained, and they tend to be less amenable to remediation later in
life. For instance, babies born with very low birth weight face higher
risks of infant mortality and long-term developmental deficits that
cannot be fully reversed \citep{cook2015understanding}. Similarly, a
child who misses critical early vaccinations is more vulnerable to
infectious diseases that can kill or disable them, which results to
lifelong negative health impacts \citep{costa2024child}. Such early-life
health setbacks may permanently impair cognitive development and
physical growth. This limits long-run productivity regardless of
subsequent improvements in schooling or income.

@@

I use this natural experiment to estimate the causal impact of increased
family size on early-childhood health outcomes. The analysis uses
nationally representative data from multiple rounds of the Philippine
Demographic and Health Survey (DHS), which include detailed information
on household structure, maternal characteristics, and child health
indicators. The primary outcomes are birth weight (measured in grams,
when reported) and whether the child received key early-life
vaccinations, including those against polio,
diphtheria--pertussis--tetanus (DPT), and measles, as a measure of
essential postnatal health investment. These outcomes are widely
recognized indicators of child well-being and reflect both biological
and behavioral dimensions of parental investment in early development.
The empirical strategy follows a difference-in-differences design that
compares child outcomes in Manila and comparable urban areas before and
after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the
absence of the contraceptive ban, trends in child health outcomes in
Manila would have evolved in parallel with those in the comparison
cities. Second, no other policy changes or economic shocks in Manila
during the study period coincided with the timing and scope of the
contraceptive policy. I probe these assumptions using falsification
tests, placebo comparisons, and robustness checks that include
city-specific time trends, mother fixed effects, and controls for
baseline demographic differences. The analysis proceeds in three stages.
I first replicate existing work \citep{dumas2019sex} to confirm that the
contraceptive ban led to an increase in fertility among affected women.
Next, I estimate the reduced-form effects of exposure to the policy on
the selected child health outcomes (birth weight, and vaccination
uptake). Finally, I examine heterogeneity in these effects across
subgroups defined by maternal education, household wealth, and access to
prenatal care. These characteristics proxy for a household's resources
and its capacity to buffer the impact of increased family size on child
health.

@@

In what follows, I review the theoretical foundations of the Q-Q model
and its evolution in the literature. I begin with the static models of
Becker \citep{becker1960economic} and Becker--Lewis
\citep{becker1973interaction}, which first formalized the trade-off
within a household utility maximization framework. I then examine
extensions to dynamic, intergenerational settings, including the
contributions of Becker and Tomes \citep{becker1976child} on child
endowments and the altruistic dynastic model associated with Barro and
Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and
unified growth models, notably Galor and Weil
\citep{galor2000population} and Galor and Moav \citep{galor2002natural},
which integrate the Q--Q mechanism into a general theory of demographic
and economic transformation.

@@

Beyond survival, early-life health indicators such as birth weight and
vaccination status are important dimensions of child quality, especially
in low-income settings. When resources are limited, parents may struggle
to meet the basic needs of each child. Empirical evidence aligns with
the quantity--quality framework: children from larger families often
have worse health outcomes at birth and are less likely to receive core
immunizations. For example, recent multi-country data show that the
share of completely unimmunized children---so-called ``zero-dose''
children---increases from 10.5\% among those with no siblings to 17.2\%
among those with four or more \citep{costa2024child}. Studies from
Southeast Asia also report that higher birth order correlates with lower
average birth weight. These patterns suggest that as fertility
increases, the health of each child may suffer due to more limited
parental investment. Studies from Southeast Asia
\citep{chen2021population} also report that higher birth order
correlates with lower average birth weight.

@@

The results are striking. The contraceptives ban led to a significant
rise in births and family size in Manila city relative to the control
areas. Correspondingly, children born in Manila during the ban era
experienced a sizable decline in educational attainment -- clear
evidence of a Q--Q trade-off precipitated by the shock. In the authors'
words, the policy-driven increase in family size ``provide{[}s{]}
evidence of a quality--quantity trade-off'': larger families, forced by
the ban, resulted in lower schooling outcomes per child.

@@

This study builds upon the foundational framework of the
quantity-quality (Q--Q) trade-off, first articulated in
\citet{becker1960economic} and later refined in
\citet{becker1973interaction}. In this model, fertility and child
quality are jointly determined under a full-income constraint, such that
increases in the number of children---without a corresponding expansion
in household resources---necessitate a reduction in the average
investment per child. While empirical applications of this framework
have traditionally focused on educational outcomes, child quality can
also be expressed in terms of early-life health, including indicators
that emerge at birth or shortly after, such as birth weight, and uptake
of essential vaccinations.

@@

In the presence of a fertility shock, holding \(I\) constant, the
average \(\kappa_j\) must fall unless offset by changes in parental
behavior or external transfers. If these inputs decline and are critical
to early-life development, observable deficits may arise in birth
weight, and children may be less likely to receive complete immunization
schedules.

The first hypothesis derived from this framework is that an exogenous
increase in fertility leads to a decline in per-child health investment,
particularly in low-income urban households with little financial slack.
A second, related hypothesis posits that this decline results in
measurable early-life health disadvantages---specifically, lower birth
weight and reduced uptake of key childhood vaccinations. These
hypotheses describe a causal pathway through which fertility shocks may
contribute to early deprivation and potentially reinforce
intergenerational cycles of poor health and limited human capital.

@@

This study draws on two primary data sources: the Philippine Demographic
and Health Surveys (DHS) and the Integrated Public Use Microdata Series
(IPUMS) census data. The DHS provides detailed child- and
household-level information on fertility behavior, maternal
characteristics, and early childhood health outcomes. I use three rounds
of the DHS---1998, 2003, and 2008---which are the only Philippine DHS
waves with publicly available GPS cluster data. These data are used to
classify respondents into treatment and control groups. Because
city-level identifiers are not included in the public-use DHS files, I
rely on approximate geographic assignment using cluster coordinates.
Clusters located within or near the administrative boundaries of the
City of Manila are classified as treated, while clusters located
elsewhere in the National Capital Region (NCR) serve as the control
group.

The DHS analytic sample is limited to children under five years of age
who were living with their mothers at the time of the survey. I examine
three core early-life health outcomes: (1) reported birth weight (in
grams); (2) maternal perception of birth size; and (3) vaccination
uptake, measured as the receipt of key early childhood immunizations
(BCG, DPT, polio, and measles). These outcomes are consistently
available across all three survey rounds and are commonly used as
proxies for child quality in studies of the quantity--quality trade-off.

To complement the DHS analysis and support the identification strategy,
I construct a panel of individual-level microdata from the IPUMS
harmonized Philippine censuses (1990, 2000, and 2010). These nationally
representative surveys provide a snapshot of women of reproductive age
(15--49) and contain detailed fertility histories---including total
children ever born, births in the past year, and number of surviving
children. I limit the sample to women ages 15--49 and construct a
composite fertility index combining several measures: total children
ever born, births in the past year, number of children under age five,
and the age of the youngest child. By comparing Manila with other
provinces and municipalities in NCR, I test whether long-run fertility
patterns reflect the contraceptive policy shock.

The 1998 DHS serves as the primary pre-policy baseline, capturing births
prior to the issuance of Executive Order No.~003. The 2008 DHS is the
main post-policy round, while the 2003 DHS is treated as a transition
period. I exclude the 1993 DHS due to the absence of geospatial
identifiers. In contrast, the census years 1990 and 2000 bracket the
policy's onset, while the 2010 census provides post-policy fertility
outcomes. This combination of sources enables both short-run and
long-run assessments of the policy's impact on fertility and child
investment.

To validate the policy shock, I use DHS region- and city-level
contraceptive use indicators, which confirm a sharp and sustained
decline in modern contraceptive use in Manila relative to other NCR
cities. These patterns reinforce the plausibility of Executive Order
No.~003 as an exogenous shock to fertility preferences and access. I
also considered the Philippine National Nutrition Survey (NNS) as an
auxiliary source, but its limited pre-2003 coverage restricts its
utility for identifying pre-policy trends. For this reason, the combined
DHS and IPUMS datasets offer the most comprehensive and temporally
balanced data available for the study's objectives.

@@

The primary outcome variables reflect early-life health and investment.
These include: (1) birth weight in grams, as recorded by the mother when
available; and (2) indicators for whether the child received core
early-life vaccinations, including BCG, DPT1--3, polio doses, and
measles.

The main explanatory variable is sibship size, defined as the total
number of surviving children ever born to the mother at the time of the
child's birth, as reported in the birth history module. This variable
captures the resource dilution mechanism central to the
quantity--quality trade-off. To isolate exogenous variation in sibship
size, I construct an instrument for fertility based on policy exposure:
an interaction of a Manila indicator and a post-ban cohort indicator.
Specifically, children born in the City of Manila in or after 2000 are
coded as treated. In extended specifications, I interact this treatment
indicator with maternal age at birth to allow for heterogeneity in
fertility responses across age cohorts, considering the fact that older
women likely exhibited lower responsiveness due to reduced fecundity.

Control variables include standard child-, maternal-, and
household-level characteristics. At the child level, I control for sex,
birth order, multiple birth status, and the preceding birth interval in
months. Maternal covariates include age at birth, educational attainment
(none, primary, secondary, higher), literacy, employment status, and age
at first birth. Household-level controls include household size,
urban/rural residence (restricted to NCR in the main analysis), and the
DHS wealth index. All regressions include survey year fixed effects to
control for secular trends and common shocks. Estimates are weighted
using DHS sample weights, and standard errors are clustered at the
primary sampling unit (PSU) level.

@@

To estimate the causal effect of fertility on early-life health
outcomes, I implement a two-stage least squares (2SLS) strategy. The
first stage is estimated at the maternal level, as fertility is a
cumulative outcome of the mother rather than the child. Importantly,
this stage also serves to establish whether the contraceptive supply ban
had a measurable impact on maternal fertility behavior. Specifically, I
use two measures of fertility: the total number of children ever born
and the number of living children at the time of survey. These outcomes
capture distinct dimensions of fertility---cumulative births and net
surviving offspring---and are regressed on an indicator for policy
exposure, defined as the interaction between a Manila municipality
indicator and a post-ban birth cohort.

In the second stage, I shift to the child level and use the number of
surviving older siblings at the time of the index child's birth as the
endogenous regressor, instrumented by policy exposure. This variable
captures sibship size at birth, and allows me to estimate the effect of
increased family size---induced by the contraception ban---on early-life
health outcomes such as birth weight, and number of vaccines received.
This two-stage framework allows me to isolate plausibly exogenous
variation in family size induced by the contraceptive supply ban in
Manila, while addressing endogeneity concerns that arise when fertility
is influenced by unobserved preferences or constraints. I estimate all
models using linear probability models (LPMs) for consistency with the
2SLS framework. All specifications apply DHS sampling weights and
cluster standard errors at the PSU level.

\section{Results and Discussion}

\subsection{Descriptive Statistics}

The following tables present summary statistics for key child health
outcomes, disaggregated by location and policy period. Table 1 presents
summary statistics for mean birth weight by location and policy period.

\input{../data-analysis/outputs/dhs_summary_birth_weight_t.tex}

Among children born in Manila, the average birth weight declined from
3,023.1 grams (SD = 712.6) in the pre-policy period to 2,969.8 grams (SD
= 619.2) after the policy's implementation. A similar decline is
observed in non-National Capital Region (NCR) areas, where mean birth
weight dropped from 3,106.4 grams (SD = 726.9) to 2,994.5 grams (SD =
681.7). In contrast, birth weights in other NCR cities remained
relatively stable over time, increasing slightly from 2,830.4 grams (SD
= 743.6) to 2,965.9 grams (SD = 793.7). The majority of births in the
sample occurred in non-NCR areas, which is expected given that these
include the rest of the country beyond Metro Manila. Notably, the
decline in birth weight was more pronounced in Manila compared to other
cities within the National Capital Region (NCR), suggesting a localized
impact of the policy. While non-NCR areas also experienced reductions,
the contrast between Manila and its neighboring NCR cities---where the
policy was not implemented---is a more relevant comparison for isolating
potential policy effects.

Lastly, Table 3 presents the distribution of childhood vaccine uptake
across policy periods and geographic areas.

\input{../data-analysis/outputs/dhs_summary_vaccines_t.tex}

Table 3 presents the distribution of childhood vaccine uptake across
policy periods and geographic areas. A substantial share of children in
all groups received no vaccines, especially in Manila (79.4\%
pre-policy, 75.2\% post-policy). However, there was a notable
post-policy increase in children receiving 5 or more vaccines in both
NCR (from 25.2\% to 38.2\%) and non-NCR cities (from 25.1\% to 33.5\%).
These shifts indicate modest improvements in vaccine coverage in areas
outside Manila, which---being unaffected by the contraceptive access
restrictions---merit further investigation into other potential drivers,
such as enhancements in health system capacity or increased
accessibility of immunization services.
